---
title: "Plot alpha diversity indices in a boxplot"
output: github_document
---
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
To compare different alpha diversity indices, we calculated species richness, Simpson index, Shannon index as well as eveness for each microhabitat per sampling period

## Load data


```{r Load Data, warning=FALSE, message=FALSE}


rm(list = ls())

library(ggplot2)
library(vegan)
library(RColorBrewer)
library(reshape2)
library(agricolae)


#setwd("03_Alpha_Diversity")


OTU_Table = as.data.frame(read.csv2("../00_Data/05_Cercozoa_Seasonal_OTU_Table_min-freq-7633_transposed_withMetadata.csv",header = T, fileEncoding="UTF-8-BOM"))
SampleMetadata = OTU_Table[,1:17]
OTU_Table = OTU_Table[,18:ncol(OTU_Table)]

```

## Calculate Alpha Diversity Indices
To calculate the diversity indices, simply load the table and caclulate the species richness per microhabitat and sampling period, then convert it into a dataframe and add all other diversity `indices` - in this case: The Simpson index, Shannon Index and eveness. Finally add the metadata and group.

```{r Calculate alpha}

richness = specnumber(OTU_Table)
df = as.data.frame(richness)


df$simpson = diversity(OTU_Table, index = "simpson")
df$shannon = diversity(OTU_Table, index = "shannon")
df$eveness = df$shannon/log(df$richness)


rownames(df) = SampleMetadata$SampleID

df$Microhabitat = SampleMetadata$Microhabitat
df$TreeSpecies = SampleMetadata$TreeSpecies
df$Season = SampleMetadata$Season
df$SamplingDescription = SampleMetadata$SamplingDescription
df$MicrohabitatPerSampling = SampleMetadata$MicrohabitatPerSampling
df$SamplingDescription = factor(df$SamplingDescription, levels = c('Autumn 2017','Spring 2018','Autumn 2018', 'Spring 2019' ))


df_melted = melt(df)

```
## Plot the Figure

Now we want to visualise the alpha diversity indices in a combined boxplot:

```{r CercozoaAlphaBoxPlot, fig.width=16, fig.height=9, dpi=600}

g = ggplot(df_melted, aes(x = Microhabitat, y = value, fill = Microhabitat)) + 
  stat_boxplot(geom = "errorbar", width = 0.1, show.legend = F) +
  geom_boxplot(show.legend = F) + 
  scale_fill_manual(values = c("#138F6A","#FF944D","#FFE4B5","#BF6692","#A8ACFF","#4363d8","#8B8989","#9A6324","#800000"), 
                    limits = c("Fresh Leaves","Bark","Arboreal Soil", "Deadwood","Hypnum","Orthotrichum","Lichen","Leaf Litter", "Soil")) + 
  scale_x_discrete(limits = c("Fresh Leaves","Deadwood","Lichen","Bark","Orthotrichum","Hypnum","Arboreal Soil","Leaf Litter", "Soil")) + 
  theme_minimal() + 
  labs(y = "Alpha Diversity", 
       x = "Microhabitat")+ 
  theme(axis.text=element_text(size=14, face = "bold"), 
        axis.title=element_text(size=16, face = "bold"), 
        plot.subtitle = element_text(size = 16, hjust = 0.5), 
        strip.text = element_text(size = 16, face = "bold"),
        axis.text.x = element_text(angle=45, hjust = 1))+
 facet_grid(variable ~ SamplingDescription, scales = "free", switch = "y")+
 theme(plot.margin = margin(1,1,1,1, "cm")) 

g
```



```{r, fig.width=16, fig.height=9, dpi=600}

#write.csv(df_melted, file = "AlphaDiv_Values.csv")



#ggsave("AlphaDiv_300_336x252.jpeg", plot = g, 
#       device = "jpeg", dpi = 300, width = 336, height = 252, 
#       units = "mm")
#ggsave("AlphaDiv_600_336x252.pdf", plot = g, 
#       device = "pdf", dpi = 600, width = 336, height = 252, 
#       units = "mm")
#ggsave("AlphaDiv_600_336x252.tif", plot = g, 
#       device = "tiff", dpi = 600, width = 336, height = 252, 
#       units = "mm")
#ggsave("AlphaDiv_600_336x252.png", plot = g, 
#      device = "png", dpi = 600, width = 336, height = 252, 
#      units = "mm")
```