---
title: "Nonmetric Multidimensional Scaling"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

NMDS is a method for visualising differences in community composition among samples, using distance matrices like the Bray-Curtis matrix. The idea is to visualise beta Diversity patterns within multiple dimensions in a 2-dimensional plot.

For more information, check out the tutorial by **[Jon Lefcheck](R.https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/)**, which guides you through the basics of an NMDS analysis in R.

## Load data


```{r Load Data, warning=FALSE, message=FALSE}

rm(list = ls())

library(vegan)
library(ggplot2)
library(funrar)
library(ggpubr)
library(RColorBrewer)
library(dplyr)


OTU_Table = as.data.frame(read.csv2("../00_Data/05_Cercozoa_Seasonal_OTU_Table_min-freq-7633_transposed_withMetadata.csv",header = T))

species = OTU_Table[,18:ncol(OTU_Table)]
species_mat = as.matrix(species)
species_mat = make_relative(species_mat)
SampleMetadata = as.data.frame(OTU_Table[,1:17])

```

## Data Preparation

For NMDS analysis we first have to cluster our sampled microhabitats by sampling perio, calculate the stress values and get the data scores. First, we need a distance matrix. We use the *vegdist* function, to build distance matrix based on Bray-Curtis-Distances. This matrix is then parsed into the *metaMDS* function. The resulting object contains a lot of information, but we need to add the metadata information for the plot.

```{r Cluster by Period, warning=FALSE, message=FALSE, results="hide"}

# group by sampling period
species_mat_Autumn2017 = species_mat[OTU_Table$SamplingDescription == "Autumn 2017",]
species_mat_Spring2018 = species_mat[OTU_Table$SamplingDescription == "Spring 2018",]
species_mat_Autumn2018 = species_mat[OTU_Table$SamplingDescription == "Autumn 2018",]
species_mat_Spring2019 = species_mat[OTU_Table$SamplingDescription == "Spring 2019",]

# calculate Bray-Curtis-distances and stress values for each sampling period

Dist_Autumn2017 = vegdist(species_mat_Autumn2017, diag = T, na.rm = T)
OTU.NMDS.bray_Autumn2017 = metaMDS(Dist_Autumn2017, k=3, trymax=300, wascores=TRUE, trace=TRUE, zerodist="add")

Dist_Spring2018 = vegdist(species_mat_Spring2018, diag = T, na.rm = T)
OTU.NMDS.bray_Spring2018 = metaMDS(Dist_Spring2018, k=3, trymax=300, wascores=TRUE, trace=TRUE, zerodist="add")


Dist_Autumn2018 = vegdist(species_mat_Autumn2018, diag = T, na.rm = T)
OTU.NMDS.bray_Autumn2018 = metaMDS(Dist_Autumn2018, k=3, trymax=300, wascores=TRUE, trace=TRUE, zerodist="add")

Dist_Spring2019 = vegdist(species_mat_Spring2019, diag = T, na.rm = T)
OTU.NMDS.bray_Spring2019 = metaMDS(Dist_Spring2019, k=3, trymax=300, wascores=TRUE, trace=TRUE, zerodist="add")

# get data scores
data.scores_sep = as.data.frame(rbind(scores(OTU.NMDS.bray_Spring2018), 
                                      scores(OTU.NMDS.bray_Spring2019),
                                      scores(OTU.NMDS.bray_Autumn2017), 
                                      scores(OTU.NMDS.bray_Autumn2018)))


rownames(data.scores_sep) = SampleMetadata$SampleID
data.scores_sep$site = rownames(data.scores_sep)
data.scores_sep$Microhabitat = SampleMetadata$Microhabitat
data.scores_sep$Stratum = SampleMetadata$Stratum
data.scores_sep$TreeSpecies = SampleMetadata$TreeSpecies
data.scores_sep$MicrohabitatPerSampling = SampleMetadata$MicrohabitatPerSampling
data.scores_sep$SamplingDescription = SampleMetadata$SamplingDescription
data.scores_sep$SamplingDate = SampleMetadata$SamplingDate
data.scores_sep$TreeSpeciesPerSampling = SampleMetadata$TreeSpeciesPerSampling

for(a in unique(SampleMetadata$MicrohabitatPerSampling)){
  a = get("a")
  a_short = gsub(" ", "", a)
  b = paste0("MicrohabitatGroup_sep.", a_short)
  b = get("b")
  #print(b)
  assign(b, data.scores_sep[data.scores_sep$MicrohabitatPerSampling == a,][chull(data.scores_sep[data.scores_sep$MicrohabitatPerSampling == a, c("NMDS1", "NMDS2")]), ])
}

MicrohabitatGroups_sep = do.call("rbind", lapply(ls(pattern = "MicrohabitatGroup_sep\\."),get))
MicrohabitatGroups_sep$SamplingDescription = factor(MicrohabitatGroups_sep$SamplingDescription, levels=c('Autumn 2017','Spring 2018','Autumn 2018','Spring 2019'))

for(a in unique(SampleMetadata$TreeSpeciesPerSampling)){
  a = get("a")
  a_short = gsub(" ", "", a)
  b = paste0("TreeGroup_sep.", a_short)
  b = get("b")
  #print(b)
  assign(b, data.scores_sep[data.scores_sep$TreeSpeciesPerSampling == a,][chull(data.scores_sep[data.scores_sep$TreeSpeciesPerSampling == a, c("NMDS1", "NMDS2")]), ])
}

TreeGroups_sep = do.call("rbind", lapply(ls(pattern = "TreeGroup_sep\\."),get))
TreeGroups_sep$SamplingDescription = factor(TreeGroups_sep$SamplingDescription, levels=c('Autumn 2017','Spring 2018','Autumn 2018','Spring 2019'))



```
## Plot with ggplot2

```{r Plot NMDS, fig.width=16, fig.height=9, dpi=600, warning=FALSE, message=FALSE}
g = ggplot() + 
  geom_polygon(data = MicrohabitatGroups_sep, 
               aes(x=NMDS1, y=NMDS2, group = MicrohabitatPerSampling, fill = Microhabitat), 
               alpha = 0.7) + 
  scale_fill_manual(values = c("#138F6A","#FF944D","#FFE4B5","#BF6692","#A8ACFF",
                               "#4363d8","#8B8989","#9A6324","#800000"), 
                    limits = c("Fresh Leaves","Bark","Arboreal Soil", 
                               "Deadwood","Hypnum","Orthotrichum",
                               "Lichen","Leaf Litter", "Soil")) + 
  geom_point(data = data.scores_sep, 
             aes(x = NMDS1, y = NMDS2, shape = TreeSpecies), 
             size = 3,
             color = "#5d5f66") + 
  geom_polygon(data = TreeGroups_sep, 
               aes(x=NMDS1, y=NMDS2, group = TreeSpeciesPerSampling, color = TreeSpecies),
               alpha = 0.7, fill = NA, linetype = "dashed", size = 1.5) +
  scale_color_manual(values = c("#c2b2b4", "#53687e", "#6b4e71"), 
                     labels = c(expression(italic("Fraxinus excelsior")), 
                               expression(italic("Quercus robur")), 
                               expression(italic("Tilia cordata")))) +
  scale_shape_manual(values = c(16,17,15), 
                     labels = c(expression(italic("Fraxinus excelsior")), 
                                expression(italic("Quercus robur")), 
                                expression(italic("Tilia cordata")))) +
  coord_equal() + 
  theme_minimal() +
  theme(axis.text=element_text(size=14), 
        axis.title=element_text(size=16, face = "bold"), 
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16, face = "bold"), 
        legend.position = "right",
        legend.text.align = 0,
        legend.direction = "vertical", 
        strip.text.y = element_text(size=16, face = "bold"), 
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)) + 
  facet_grid(cols = vars(SamplingDescription))+
  facet_wrap(~SamplingDescription, ncol=2)+
  theme(strip.text.x = element_text(size=16,face="bold"))+
  theme(plot.margin = margin(1,1,1,1, "cm"))

g

```

```{r, fig.width=16, fig.height=9, dpi=600, echo=FALSE}
#ggsave("NMDS_300_336x252.jpeg", plot = g, 
#        device = "jpeg", dpi = 300, width = 336, height = 252, 
#        units = "mm")
#ggsave("NMDS_600_336x252.pdf", plot = g, 
#       device = "pdf", dpi = 600, width = 336, height = 252, 
#       units = "mm")
#ggsave("NMDS_600_336x252.tif", plot = g, 
#       device = "tiff", dpi = 600, width = 336, height = 252, 
#       units = "mm")
#ggsave("NMDS_600_336x252.png", plot = g, 
#       device = "png", dpi = 600, width = 336, height = 252, 
#       units = "mm")
```
