---
title: "Principle Coordinates Analysis"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For Analysis of beta diversity across communities detected on fresh leaves a Principle Coordinates Analysis was performed. The PCoA  is a method to explore and to visualize similarities or dissimilarities of data. It starts with a similarity matrix or dissimilarity matrix (= distance matrix) and assigns for each item a location in a low-dimensional space.

## Load data


```{r Load Data, warning=FALSE, message=FALSE}
rm(list = ls()) 

library(vegan)
library(ggplot2)
library(plyr)
library(funrar) 
library(ggpubr)

OTU_Table = as.data.frame(read.csv2("../00_Data/05_Cercozoa_Seasonal_OTU_Table_min-freq-7633_transposed_withMetadata.csv",header = T))


species = OTU_Table[,18:ncol(OTU_Table)]
species.mat = as.matrix(species)
species.mat = make_relative(species.mat)
SampleMetadata = as.data.frame(OTU_Table[,1:17])

rownames(species.mat) <- OTU_Table[,1]

```

## Data Preparation

First step is to calculate a distance matrix. The most popular function to perform a classical scaling is *cmdscale* (which comes with the default distribution of R). Here we use Bray-Curtis distance metric. Next we have to check the caculated eigenvalues and interpret the observed variation. Finally we only have to group our data by "sampling period" for visualization.

```{r PCoA Microhabitats, warning=FALSE, message=FALSE, results="hide"}

# calculate Bray-Curtis distances
pcoa <- cmdscale(vegdist(species.mat, "bray"), eig=TRUE, k=2) 

# explained variation by axis 1 and axis 2 of PCoA
explainedvar1<-round(pcoa$eig[1]/sum(pcoa$eig),2)*100
explainedvar2<-round(pcoa$eig[2]/sum(pcoa$eig),2)*100

#explainedvar1 #  the first x-axis explains 25% of variation 
#explainedvar2 # the second y-axis explains 11% of variation 


pcoa<-as.data.frame(cbind(pcoa$points, pcoa$eig)) # keep only the first 3 columns as data frame
#dim(pcoa)

# group by sampling period
pcoaF18<-as.data.frame(pcoa[c(1:65),])
pcoaF19<-as.data.frame(pcoa[c(66:141),])
pcoaH17<-as.data.frame(pcoa[c(142:221),])
pcoaH18<-as.data.frame(pcoa[c(222:290),])


# create a column called "Season"
pcoaF18$Season=rep("Spring 2018",nrow(pcoaF18))
pcoaF19$Season=rep("Spring 2019",nrow(pcoaF19))
pcoaH17$Season=rep("Autumn 2017",nrow(pcoaH17))
pcoaH18$Season=rep("Autumn 2018",nrow(pcoaH18))

pcoabyseason<-rbind(pcoaH17, pcoaF18, pcoaH18, pcoaF19)

#  define function and calculate the points for the PCoA by applying the function to visualize data
find_hull.pcoa <- function(df) df[chull(df$V1, df$V2), ]

hulls.pcoaSe <- ddply(pcoabyseason,"Season", find_hull.pcoa)


```

##ggplot 

```{r Plot All Samples,warning=FALSE, message=FALSE}


PCOA_All <- ggplot()+
  geom_point(data=pcoabyseason, aes(x=V1,y=V2, col=Season,
                                    shape=Season), size=3)+
  geom_polygon(data=hulls.pcoaSe,aes(x=V1, y=V2, group = Season, fill = Season),alpha=0.2)+
  scale_shape_manual(values=c(16,16,17,17))+
  theme_minimal()+
  scale_fill_manual(values = c("#094d35","#660033","#B5A642","#1c9099"), 
                    limits = c("Autumn 2017","Autumn 2018","Spring 2018","Spring 2019")) +  
  scale_color_manual(values=c("#094d35","#660033","#B5A642","#1c9099"))+
  xlab('PCoA1')+
  ylab('PCoA2')+
  theme(axis.text=element_text(size=14), 
        axis.title=element_text(size=16, face = "bold"), 
        legend.text = element_text(size = 16), 
        legend.title = element_blank(),
        legend.position = "bottom", 
        legend.direction = "horizontal", 
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5))+
  labs(y= "PCoA2 - 11%", x = "PCoA1 - 25%")
        
PCOA_All

```
Repeat every step for fresh leaves only.

```{r PCoA Leaves, warning=FALSE, message=FALSE, results="hide"}


# repeat for leaves only
OTU_Table_Leaves <- OTU_Table[ which(OTU_Table$Microhabitat == 'Fresh Leaves'),]


species.leaves <- OTU_Table_Leaves[,18:ncol(OTU_Table)]
species.leaves.mat = as.matrix(species.leaves)
species.leaves.mat = make_relative(species.leaves.mat)
#dim(species.leaves)
 
rownames(species.leaves.mat) <- OTU_Table_Leaves[,1]
 

#calculate PCoA 
pcoa <- cmdscale(vegdist(species.leaves.mat, "bray"), eig=TRUE, k=2) 


# explained variation by axis 1 and axis 2 of PCoA
explainedvar1<-round(pcoa$eig[1]/sum(pcoa$eig),2)*100
explainedvar2<-round(pcoa$eig[2]/sum(pcoa$eig),2)*100

#explainedvar1 #  the first x-axis explains 21% of variation 
#explainedvar2 # the second y-axis explains 17% of variation 

pcoa<-as.data.frame(cbind(pcoa$points, pcoa$eig)) # keep only the first 3 columns as data frame
#dim(pcoa)

# group by sampling period
pcoaF18<-as.data.frame(pcoa[c(1:5),])
pcoaF19<-as.data.frame(pcoa[c(6:10),])
pcoaH17<-as.data.frame(pcoa[c(11:19),])
pcoaH18<-as.data.frame(pcoa[c(20:25),])


# create a column called "Season"
pcoaF18$Season=rep("Spring 2018",nrow(pcoaF18))
pcoaF19$Season=rep("Spring 2019",nrow(pcoaF19))
pcoaH17$Season=rep("Autumn 2017",nrow(pcoaH17))
pcoaH18$Season=rep("Autumn 2018",nrow(pcoaH18))

pcoabyseason<-rbind(pcoaH17, pcoaF18, pcoaH18, pcoaF19)

# calculate the points for the PCoA by applying the function to visualize data
find_hull.pcoa <- function(df) df[chull(df$V1, df$V2), ]

hulls.pcoaSe <- ddply(pcoabyseason,"Season", find_hull.pcoa)



```

```{r Plot Leaves only, warning=FALSE, message=FALSE}

PCOA_Leaves <- ggplot()+
  geom_point(data=pcoabyseason, aes(x=V1,y=V2, col=Season,
                                    shape=Season), size=3)+
  geom_polygon(data=hulls.pcoaSe,aes(x=V1, y=V2, group = Season, fill = Season),alpha=0.2)+
  scale_shape_manual(values=c(16,16,17,17))+
  theme_minimal()+
  scale_fill_manual(values = c("#094d35","#660033","#B5A642","#1c9099"), 
                    limits = c("Autumn 2017","Autumn 2018","Spring 2018","Spring 2019")) +  
  scale_color_manual(values=c("#094d35","#660033","#B5A642","#1c9099"))+
  xlab('PCoA1')+
  ylab('PCoA2')+
  theme(axis.text=element_text(size=14), 
        axis.title=element_text(size=16, face = "bold"), 
        legend.text = element_text(size = 16), 
      
        legend.title = element_blank(),
        legend.position = "bottom", 
        legend.direction = "horizontal", 
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5))+
  labs(y= "PCoA2 - 17%", x = "PCoA1 - 21%")
        
PCOA_Leaves

```

```{r combine Plots, fig.width=16, fig.height=9, dpi=600, warning=FALSE, message=FALSE}


combi = ggarrange(PCOA_All, PCOA_Leaves,
                  labels = c("A", "B"), 
                  ncol = 2, nrow = 1, font.label = list(size = 16, color = "black"),
                  common.legend = T, legend = "bottom", align = "h")+
        theme(plot.margin = margin(1,1,1,1, "cm")) 

combi

#save plot

#ggsave("PCA_300_336x168.jpeg", plot = combi, 
#       device = "jpeg", dpi = 300, width = 336, height = 168, 
#       units = "mm")
#ggsave("PCA_600_336x168.pdf", plot = combi, 
#       device = "pdf", dpi = 600, width = 336, height = 168, 
#       units = "mm")
#ggsave("PCA_600_336x168.tif", plot = combi, 
#       device = "tiff", dpi = 600, width = 336, height = 168, 
#       units = "mm")
#ggsave("PCA_600_336x168.png", plot = combi, 
#       device = "png", dpi = 600, width = 336, height = 168, 
#       units = "mm")
```
