---
title: "Variation Partitioning Analysis"
output: github_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In the following section we want to calculate the proportion of variation in species composition explained by each variable.Variation partitioning is a method of choice for the interpretation of beta diversity using tables of environmental and spatial variables. The technique of variation partitioning is used when two or more complementary sets of hypotheses can be invoked to explain the variation of an ecological response variable.
A detailed description of variation  partitioning can be found [here](https://academic.oup.com/jpe/article/1/1/3/1130269).

##Load data

```{r Load Data, warning=FALSE, message=FALSE}
rm(list = ls()) 

library(vegan)
library(plyr)
library(funrar) 
library(ggplot2)
library(ggforce)
library(ggpubr)


OTU_Table = as.data.frame(read.csv2("../00_Data/05_Cercozoa_Seasonal_OTU_Table_min-freq-7633_transposed_withMetadata.csv",header = T))

#All
species = OTU_Table[,18:ncol(OTU_Table)]
species.mat = as.matrix(species)
species.mat = make_relative(species.mat)
SampleMetadata = as.data.frame(OTU_Table[,1:17])

#We need the Hellinger-transformed OTUs table with OTUs as columns and a table of metadata: Microhabitat, season and treespecies.
OTUCer.hel = decostand(species.mat, "hellinger")


#Ground
OTU_Ground <- OTU_Table[ which(OTU_Table$Stratum=='Ground'), ]

species.Ground = OTU_Ground[,18:ncol(OTU_Ground)]
species.mat.Ground = as.matrix(species.Ground)
species.mat.Ground = make_relative(species.mat.Ground)
SampleMetadata.Ground = as.data.frame(OTU_Ground[,1:17])

OTUCer.hel.Ground = decostand(species.mat.Ground, "hellinger")

#Canopy
OTU_Canopy <- OTU_Table[ which(OTU_Table$Stratum=='Canopy'), ]

species.Canopy = OTU_Canopy[,18:ncol(OTU_Canopy)]
species.mat.Canopy = as.matrix(species.Canopy)
species.mat.Canopy = make_relative(species.mat.Canopy)
SampleMetadata.Canopy = as.data.frame(OTU_Canopy[,1:17])

OTUCer.hel.Canopy = decostand(species.mat.Canopy, "hellinger")


```
##Calulate Variation Partitioning

First we plot the variation partitioning of our variables with the `varprat` function within the *vegan* package. 
Variation partitioning (using RDA, CCA or db-RDA) among up to four matrices of environmental variables. First argument (Y) is dependent variable, usually the matrix of species composition (the function calculates RDA, or, if chisquare = TRUE, CCA), but could be also only a single variable (in that case it calculates linear regression) or distance matrix (applying db-RDA using the function capscale). Further arguments (up to four) are (groups of) explanatory variables. The function uses either formula interface (with ~, see examples) or matrices. 

```{r Calculate Varprat, warning=FALSE, message=FALSE, results="hide"}
#All
Cervarp.all <- varpart(OTUCer.hel, ~ Season, ~ Microhabitat, ~TreeSpecies, data = SampleMetadata)
#Ground
Cervarp.Ground <- varpart(OTUCer.hel.Ground, ~ Season, ~ Microhabitat, ~TreeSpecies, data = SampleMetadata.Ground)
#Canopy
Cervarp.Canopy <- varpart(OTUCer.hel.Canopy, ~ Season, ~ Microhabitat, ~TreeSpecies, data = SampleMetadata.Canopy)

```

##Plot Variation Partitioning 


```{r Plot Varprat, fig.width=5, fig.height=5, units=cm, dpi=100, warning=FALSE, message=FALSE,}
#All
plot(Cervarp.all, digits = 1, Xnames = c('Season', 'Microhabitat', 'TreeSpecies'), bg = c('navy', 'tomato', 'green'))
#Ground
plot(Cervarp.Ground, digits =1, Xnames = c('Season', 'Microhabitat', 'TreeSpecies'), bg = c('navy', 'tomato', 'green'))
#Canopy
plot(Cervarp.Canopy, digits = 1, Xnames = c('Season', 'Microhabitat', 'TreeSpecies'), bg = c('navy', 'tomato', 'green'))
```

##Plot with ggplot
Unfortunately it is not so easy to modify the plot afterwards. Consequently, we repeat the plotting with a dummy dataframe in *ggplot* to modify coloration and save settings.

```{r Ggplot Varprat, fig.width=3, fig.height=3, units=cm, dpi=100, warning=FALSE, message=FALSE}
#create dummy dataframe
df.venn.all <- data.frame(x = c(3, 1, 2),y = c(1, 1,2.8),labels = c('Microhabitat\n0.32', 'Season\n0.01',"Tree species\n0.03"))
#Ground 
df.venn.Ground<- data.frame(x = c(3, 1, 2),y = c(1, 1,2.8),labels = c('Microhabitat\n0.31 **', 'Season\n0.05 **',"Tree species\n0.02 **"))
#Canopy
df.venn.Canopy<- data.frame(x = c(3, 1, 2),y = c(1, 1,2.8),labels = c('Microhabitat\n0.18 **', 'Season\n0.02 **',"Tree species\n0.05 **"))

#Plot
#All
Varpart <- ggplot(df.venn.all, aes(x0 = x, y0 = y, r = 1.5, fill = labels)) +
                  geom_circle(alpha = .6, size = 1, colour = "#3f3a3a",show.legend = FALSE) +
                  coord_fixed()+
                  annotate("text", x = df.venn.all$x , y = df.venn.all$y,label=df.venn.all$labels ,size = 6) +
                  scale_fill_manual(values = c("#660033","#B5A642","#094d35")) +
                  theme_void()
                  #annotate("text", x = 2 , y =1,label="A and B" ,size = 4) +
                  #annotate("text", x = 1.35 , y =2,label="B and C" ,size = 4) +
                  #annotate("text", x = 2.7 , y =2,label="A and C" ,size = 4) +
                  #annotate("text", x = 2 , y =1.6,label="A and B and C" ,size = 2)+theme_void()

Varpart

#Ground
A <- ggplot(df.venn.Ground, aes(x0 = x, y0 = y, r = 1.5, fill = labels)) +
                  geom_circle(alpha = .6, size = 1, colour = "#3f3a3a",show.legend = FALSE) +
                  coord_fixed()+
                  annotate("text", x = df.venn.Ground$x , y = df.venn.Ground$y,label=df.venn.Ground$labels ,size = 6) +
                  scale_fill_manual(values = c("#82662a","#bea773","#daba73")) +
                  theme_void()+
                  annotate("text", x = 3.5 , y =-0.8,label="Residuals: 0.65" ,size = 4.5) #+
                  #annotate("text", x = 2 , y =1,label="A and B" ,size = 4) +
                  #annotate("text", x = 1.35 , y =2,label="B and C" ,size = 4) +
                  #annotate("text", x = 2.7 , y =2,label="A and C" ,size = 4) +
                  #annotate("text", x = 2 , y =1.6,label="A and B and C" ,size = 2)+theme_void()

A

#Canopy

B <- ggplot(df.venn.Canopy, aes(x0 = x, y0 = y, r = 1.5, fill = labels)) +
                  geom_circle(alpha = .6, size = 1, colour = "#3f3a3a",show.legend = FALSE) +
                  coord_fixed()+
                  annotate("text", x = df.venn.Canopy$x , y = df.venn.Canopy$y,label=df.venn.Canopy$labels ,size = 6) +
                  scale_fill_manual(values = c("#42614a","#61a072","#8fd7a2")) +
                  theme_void()+
                  annotate("text", x = 3.5 , y =-0.8,label="Residuals: 0.76" ,size = 4.5)
                  #annotate("text", x = 2 , y =1,label="A and B" ,size = 4) +
                  #annotate("text", x = 1.35 , y =2,label="B and C" ,size = 4) +
                  #annotate("text", x = 2.7 , y =2,label="A and C" ,size = 4) +
                  #annotate("text", x = 2 , y =1.6,label="A and B and C" ,size = 2)+theme_void()

B

```
Combine plots:
```{r combine Plots, fig.width=16, fig.height=9, dpi=600, warning=FALSE, message=FALSE}


combi = ggarrange(A, B,
                  labels = c("A", "B"), 
                  ncol = 2, nrow = 1, font.label = list(size = 16, color = "black"),
                  common.legend = T, legend = "bottom", align = "h")+
        theme(plot.margin = margin(1,1,1,1, "cm")) 

combi

#save plot
#ggsave("VarPar_300_336x168.jpeg", plot = combi, 
#       device = "jpeg", dpi = 300, width = 336, height = 168, 
#       units = "mm")
#ggsave("VarPar_600_336x168.pdf", plot = combi, 
#       device = "pdf", dpi = 600, width = 336, height = 168, 
#       units = "mm")
#ggsave("VarPar_600_336x168.tif", plot = combi, 
#       device = "tiff", dpi = 600, width = 336, height = 168, 
#       units = "mm")
#ggsave("VarPar_600_336x168.png", plot = combi, 
#       device = "png", dpi = 600, width = 336, height = 168, 
#       units = "mm")
```

##Variation Partitioning Analysis - Forward Selection
To identify the explanatory variables that significantly explained variation in protistan communities, forward selection was performed using the `ordistep` function within the *vegan* package (1000 Monte Carlo permutations, alpha < 0.05). .


```{r Variation Partitioning, warning=FALSE, message=FALSE, results="hide"}

#Statistics: forward selection

#Ground
rda.all<-rda(OTUCer.hel.Ground~Season+Microhabitat+TreeSpecies, data=SampleMetadata.Ground)
rda.all


# Forward selection using vegan???s ordistep().
# *******************************************
# ->Hint: This function allows the use of factors. Options are also available for stepwise and backward selection of the explanatory variables.
step.forward <- ordistep(rda(OTUCer.hel.Ground ~ 1, data=SampleMetadata.Ground), scope=formula (rda.all), direction="forward", pstep=1000)

#Start: OTUCer.hel.Ground ~ 1 

#               Df     AIC       F Pr(>F)   
#+ Microhabitat  1 -76.004 26.1939  0.005 **
#+ Season        1 -56.952  3.4393  0.010 **
#+ TreeSpecies   2 -54.181  1.3060  0.160   
#---
#Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

#Step: OTUCer.hel.Ground ~ Microhabitat 

#              Df     AIC      F Pr(>F)   
#+ Season       1 -79.191 5.1488  0.005 **
#+ TreeSpecies  2 -75.755 1.8096  0.005 **
#---
#Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

#Step: OTUCer.hel.Ground ~ Microhabitat + Season 

#              Df     AIC      F Pr(>F)   
#+ TreeSpecies  2 -79.288 1.9484  0.005 **
#---
#Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

#Step: OTUCer.hel.Ground ~ Microhabitat + Season + TreeSpecies 


#Canopy
rda.all<-rda(OTUCer.hel.Canopy~Season+Microhabitat+TreeSpecies, data=SampleMetadata.Canopy)
rda.all


# Forward selection using vegan???s ordistep().
# *******************************************
# ->Hint: This function allows the use of factors. Options are also available for stepwise and backward selection of the explanatory variables.
step.forward <- ordistep(rda(OTUCer.hel.Canopy ~ 1, data=SampleMetadata.Canopy), scope=formula (rda.all), direction="forward", pstep=1000)

#Start: OTUCer.hel.Canopy ~ 1 

#               Df     AIC      F Pr(>F)   
#+ Microhabitat  6 -286.40 9.0477  0.005 **
#+ TreeSpecies   2 -256.72 6.2777  0.005 **
#+ Season        1 -251.65 5.3221  0.005 **
#---
#Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

#Step: OTUCer.hel.Canopy ~ Microhabitat 

#              Df     AIC      F Pr(>F)   
#+ TreeSpecies  2 -298.02 7.7636  0.005 **
#+ Season       1 -291.16 6.6192  0.005 **
#---
#Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

#Step: OTUCer.hel.Canopy ~ Microhabitat + TreeSpecies 

#         Df     AIC      F Pr(>F)   
#+ Season  1 -303.21 6.9828  0.005 **
#---
#Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1

#Step: OTUCer.hel.Canopy ~ Microhabitat + TreeSpecies + Season 


```


```{r Statistics, warning=FALSE, message=FALSE, results="hide", echo=FALSE}
#The interpretation the results should be based on adjusted R2, although raw R2 is also reported (for CCA, adjusted R2 is calculated by permutation method of [Peres-Neto et al. 2006](https://esajournals.onlinelibrary.wiley.com/doi/full/10.1890/0012-9658%282006%2987%5B2614%3AVPOSDM%5D2.0.CO%3B2?casa_token=tRcDjcXcbNwAAAAA%3AugcpGMzywSAYRFWifRSZJ-0fLtStFiHFuqDP1LnvoSLTdkJS_EtBhbnMCIjd0fl7nmpkonuTVIraoDafeQ) and may slightly vary between re-analyses of the same data; the argument permutations specifies the number of permutations used to calculate adjusted R2 in CCA)

#Ground
rda.all<-rda(OTUCer.hel.Ground~Season+Microhabitat+TreeSpecies, data=SampleMetadata.Ground)
rda.all
R2all <- RsquareAdj(rda.all)$adj.r.squared
R2all #0.3546248


rda.MS<-rda(OTUCer.hel.Ground~Microhabitat+Season, data=SampleMetadata.Ground)
rda.MS 
R2MS <- RsquareAdj(rda.MS)$adj.r.squared
R2MS   #  0.3342232


rda.TS<-rda(OTUCer.hel.Ground~TreeSpecies+Season, data=SampleMetadata.Ground)
rda.TS
R2TS <- RsquareAdj(rda.TS)$adj.r.squared
R2TS   #  0.03582976


rda.MT<-rda(OTUCer.hel.Ground~Microhabitat+TreeSpecies, data=SampleMetadata.Ground)
rda.MT
R2MT <- RsquareAdj(rda.MT)$adj.r.squared
R2MT   # 0.341761

#Getting values for individual fractions, S(Season), M (Microhabitat), T(Tree species), substract: 
R2S<-(R2MS+R2TS)-R2all
R2S   #   0.01542819
R2M<-R2MS-R2S
R2M   # 0.318795
R2T<-R2TS-R2S
R2T   # 0.02040156

# conditional effect of Microhabitat is highest  

#Now, when we know both simple and conditional effect of each variables, we may want to know whether these variances are significant, and hence worth of interpreting. Results from varpart contain the column testable with logical values indicating whether given fraction is testable or not. To test each of them, we will need the models defined above, and the function anova, which (if applied on single object resulting from rda or cca method, returns Monte Carlo permutation test of the predictor effect). For this, we need to first define also partial ordination models with one variable as exlanatory and the other as covariable.


#Ground
anova(rda.all <- rda (OTUCer.hel.Ground ~ Season+Microhabitat+TreeSpecies, data = SampleMetadata.Ground))
#Model: rda(formula = OTUCer.hel.Ground ~ Season + Microhabitat + TreeSpecies, data = SampleMetadata.Ground)
#         Df Variance     F Pr(>F)    
#Model     4  0.16165 9.517  0.001 ***
#Residual 58  0.24629     

#seperated
anova(rda.MS <- rda (OTUCer.hel.Ground ~ Microhabitat + Condition (Season), data = SampleMetadata.Ground))
#Model: rda(formula = OTUCer.hel.Ground ~ Microhabitat + Condition(Season), data = SampleMetadata.Ground)
#         Df Variance      F Pr(>F)    
#Model     1  0.12333 28.154  0.001 ***
#Residual 60  0.26283  

anova(rda.SM <- rda (OTUCer.hel.Ground ~ Season + Condition (TreeSpecies), data = SampleMetadata.Ground))      
#Model: rda(formula = OTUCer.hel ~ Season + Condition(Microhabitat), data = SampleMetadata)
# Df Variance      F Pr(>F)    
#Model     1 0.022555 5.1488  0.001 ***
#Residual 60 0.262832  

anova(rda.ST <- rda (OTUCer.hel.Ground ~ Season + Condition (TreeSpecies), data = SampleMetadata.Ground))
#Model: rda(formula = OTUCer.hel ~ Season + Condition(TreeSpecies), data = SampleMetadata)
#              Df Variance    F Pr(>F)   
#Model     1  0.02177 3.48  0.004 **
#Residual 59  0.36914              


anova(rda.TS <- rda (OTUCer.hel.Ground ~ TreeSpecies + Condition (Season), data = SampleMetadata.Ground))
#Model: rda(formula = OTUCer.hel ~ TreeSpecies + Condition(Season), data = SampleMetadata)
#Df Variance      F Pr(>F)
#Model     2  0.01702 1.3601  0.162
#Residual 59  0.36914 

anova(rda.TM <- rda (OTUCer.hel ~ TreeSpecies + Condition (Microhabitat), data = SampleMetadata))
#Model: rda(formula = OTUCer.hel ~ TreeSpecies + Condition(Microhabitat), data = SampleMetadata)
#          Df Variance      F Pr(>F)    
#Model      2 0.013034 7.0663  0.001 ***
#Residual 279 0.257314      

anova(rda.MT <- rda (OTUCer.hel ~ Microhabitat + Condition (TreeSpecies), data = SampleMetadata))
#Model: rda(formula = OTUCer.hel ~ Microhabitat + Condition(TreeSpecies), data = SampleMetadata)
#            Df Variance      F Pr(>F)    
#Model      8  0.13534 18.343  0.001 ***
#Residual 279  0.25731                  

       

```


