---
title: "Shared OTUs"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
To compare species richness of different sampling periods, we calculated shared OTUs.

## Load data


```{r Load Data, warning=FALSE, message=FALSE}

rm(list = ls())

library(vegan)
library(plyr)
library(ggplot2)
library(ggpubr)
library(UpSetR)
library(ggupset)
library(tidyverse)
library(cowplot)


OTU_Table = as.data.frame(read.csv2("../00_Data/05_Cercozoa_Seasonal_OTU_Table_min-freq-7633_transposed_withMetadata.csv",header = T))



SampleMetadata = OTU_Table[,1:18]
Period = SampleMetadata$SamplingDescription
OTU_Table = OTU_Table[,18:ncol(OTU_Table)]

#Aggregate by Sampling Period
Aggregated_Period = ddply(OTU_Table, "Period", numcolwise(sum))
rownames(Aggregated_Period) = Aggregated_Period$Period
Aggregated_Period = Aggregated_Period[,-1]

#make incidence based
Aggregated_Period[,] = ifelse(Aggregated_Period[,] > 0, 1, 0)
t_Aggregated_Period = as.data.frame(t(Aggregated_Period))
```
## UpSetR

Now we want to visualise shared OTUs between each sampling period using the UpSetR package.

```{r UpsetSharedOTUs, fig.width=16, fig.height=9, dpi=600, warning=FALSE, message=FALSE}

U = upset(t_Aggregated_Period, 
          nsets = 4, # Sets are the sampling periods in this case
          order.by = c("degree","freq"), # choose between frequency or degree
          decreasing = T, group.by = "sets", 
          nintersects = 6, # number of combinations to display
          empty.intersections = TRUE, 
          mainbar.y.label = "Shared OTUs", 
          sets.x.label = "Number of OTUs per Year", 
          text.scale = 1.8, set_size.show = F)

U

``` 

## Data Preparation for ggupset
As an input, we need a tibble (another version of a dataframe). To generate a tibble from a presence/absence matrix, we need to convert it with using the tidyverse package.

```{r Create tibble period, message=FALSE}


#sort data
Aggregated_Period_TF = ifelse(Aggregated_Period[,] > 0, 
                                    TRUE, FALSE)

tidy_Periods = as_tibble(Aggregated_Period_TF, 
                               rownames = "Periods") %>% 
  gather(OTU, Presence, -Periods) %>% 
  filter(Presence) %>% 
  select(- Presence)

tidy_Dataframe = tidy_Periods %>% 
  group_by(OTU) %>% 
  dplyr::summarise(shared_OTUs = list(Periods))

```
## ggupset

Now we want to visualise shared OTUs between each sampling period using the ggupset package.

```{r ggupsetSharedOTUs, fig.width=16, fig.height=9, dpi=600, warning=FALSE, message=FALSE}

a = ggplot(tidy_Dataframe, aes(x = shared_OTUs)) +
  # plot the shard OTUs
  geom_bar() +
  geom_label(aes(label=..count..), 
             stat="count", 
             position=position_stack(),
             size=5) +
  # convert to combination matrix
  scale_x_upset() +
  # display only 8 intersections
  axis_combmatrix(xlim = c(0, 8.35)) +
  theme_combmatrix(combmatrix.label.make_space = F, 
                   combmatrix.label.text = element_text(size=16, 
                                                        face = "bold"),
                   axis.text.y = element_text(size = 16), 
                   axis.title=element_text(size=16, face = "bold"), 
                   panel.background = element_blank(), 
                   panel.grid.major.y = element_line(colour = "black"), 
                   panel.grid.minor.y = element_line(color = "black"), 
                   panel.grid.major.x = element_blank(), 
                   panel.grid.minor.x = element_blank(), 
                   plot.margin = unit(c(0.5,0.5,0.5,1.8), "cm"))+ 
  labs(x = NULL, y = "Number of shared OTUs")

a

``` 

## Add a seperate Barchart to the Plot

Now we want to add the missing left bar chart of the number of OTUs.

```{r Add Barchart, fig.width=16, fig.height=9, dpi=600, warning=FALSE, message=FALSE}

tidy_Periods$Periods = factor(tidy_Periods$Periods, levels = rownames(as.data.frame(specnumber(Aggregated_Period) %>% sort())))

b = ggplot(tidy_Periods, aes(x = Periods)) + 
  geom_bar() + 
  coord_flip() +  
  scale_y_reverse() + 
  theme_minimal() + 
  labs(y = "Number of OTUs", x = NULL) + 
  theme(axis.text.y = element_blank(),
        axis.title.x = element_text(face = "bold", size = 16), 
        axis.text.x = element_text(size = 15), 
        axis.ticks.x = element_line(),
        axis.line.x = element_line(),
        axis.line.y = element_blank(), 
        panel.grid = element_blank(),
        panel.background = element_blank(), 
        plot.background = element_blank())
ab = ggdraw() + 
  draw_plot(a, x = 0.23, y = 0.06, height = 0.9, width = 0.8)+
  draw_plot(b, x = 0.03, y = 0, width = 0.2, height = 0.23)+
  theme(plot.margin = margin(1,1,1,1, "cm")) 
ab
```

```{r, fig.width=16, fig.height=9, dpi=600}
#ggsave("Upset_300_336x168.jpeg", plot = ab, 
#       device = "jpeg", dpi = 300, width = 336, height = 168, 
#       units = "mm")
#ggsave("Upset_600_336x168.pdf", plot = ab, 
#       device = "pdf", dpi = 600, width = 336, height = 168, 
#       units = "mm")
#ggsave("Upset_600_336x168.tif", plot = ab, 
#       device = "tiff", dpi = 600, width = 336, height = 168, 
#       units = "mm")
#ggsave("Upset_600_336x168.png", plot = ab, 
#       device = "png", dpi = 600, width = 336, height = 168, 
#       units = "mm")

```